Mining Interaction Log Data in a Creative Arts Practice using Macro Gestures and Transition Matrices

### 2015-09-21 flux-entropy-paper/flux-entropy-paper.tex
new descriptive data table
update length median
intro fixed up

\begin{abstract}
  In this note we describe the analysis of a data archive for an
  ensemble musical performance practice on touch screens where
  transition matrix measures are able to distinguish between sessions
  of different contexts, and the internal structure of these sessions.
  The archive includes 95 touch data logs of ensemble touch-screen
  interactions in concerts, rehearslas and formal HCI studies of
  groups of musicians over more than two years.
  The data analysis has been undertaken by first classifying these
  data into higher level gesture types and summarising the performers'
  transitions between these gestures.
  We find that two transition-matrix measure, flux, and entropy,
  can be used to distinguish different performance types and contexts,
  as well as different phases (beginning, middle and end) of all of
  the performances in our corpus.
\end{abstract}

@@

Modern computing devices and interfaces continue to present
opportunities to support new kinds of creative expression and
collaborative interaction~\cite{Resnick:2005yu}. However, it has been
acknowledged that understanding and evaluating interfaces in creative
contexts can be challenging~\cite{Shneiderman:2007qv}. In the context
of live performance, interface evaluation using questionnaires, or
ethnographic data gathering can miss the dynamic need of artists in
the moment of performance and may be impractical in live performative
contexts. When performers use a combuter-based artistic interface,
logged interaction data which has been collected for curation and
preservation~\cite{England:2014ys}, might also allow quantitative
investigation into these performances.

In this note we describe an archive of interaction data logs that has
been accumulated to support a long-term practice-led research program
into simple iPad interfaces for group music-making similar to those
described in Martin et al~\cite{Martin:2014cr}. Over a period of more
than two years (April 2013--June 2015), we collected data from group
music-making sessions where the participants used custom-built,
touch-screen apps on Apple iPad devices. Although several different
apps have been included in our corpus, they shared a common mode of
free-form touch interaction where tapping produced short sounds and
swirling or swiping produced long sounds. Apart from this commonality,
the apps featured a variety of sound palettes, different kinds of
visual feedback, and different kinds of networked interactions
(between the iPads in an ensemble and with a central server).

In all of our interaction sessions, every touch interaction---every
touch down, dragging movement and release---was time-stamped and
logged to disk for later analysis. These sessions included rehearsals
without an audience as well as live performances at arts festivals and
other events. They also included interactions during
formally-structured HCI experiments where participant compared
multiple apps and interfaces. Some of the performances were iPad-only
and others used iPads together with other acoustic percussion
instruments. Although the majority of the sessions were free-form
improvisations, a number of rehearsals and performances involved
composed pieces where the musicians read and performed percussion
notation using the apps.

Representative photographs of some of these session contexts
(rehearsal, performance including other acoustic instruments and iPad
only performance) are are shown
Figure~\ref{fig:three-performance-contexts}) and other descriptive
statistics about the sessions are shown in Table \ref{corpus-table}.
In our corpus, the median length of sessions is 7m29s and the median
number of participants per ensemble is four. Metadata was included
together with the interaction data for each session to record the
\emph{context} (performance, rehearsal, study), session \emph{type}
(improvisation or composition), and \emph{instrumentation}
(iPads-only, or iPads with acoustic percussion). The distribution of
recorded sessions by context and type is shown in Figure
\ref{fig:count-data}.

@@

% TODO fix this table to remove descriptive data about the
% demonstration sessions.
\begin{table}
\centering
\begin{tabular}{l|llll}
\hline
            & Total & Min  & Median   & Max     \\ 
\hline
N           & 150       &         &        &  \\
%Record size & 306MB     &         &        &   \\
Length      & 20H34M22S & 1M0S    & 7M29S  & 26M0S\\
Participants& 524       & 1       & 4      & 9    \\
Flux        &           & 0.02239 & 0.1445 & 0.3246\\
Entropy     &           &  &  & \\          
\hline
\end{tabular}
\caption{
  Representative data about the corpus of musical touch-screen
  interaction sessions used in this study. These sessions were
  rehearsals, performances, or studies and 
  participants improvised or performed compositions. Each record
  consisted of touch-data recorded in CSV format and was later analysed
  with a gestural classification system. Some participants were
  present in multiple sessions, however this data was not tracked in
  the corpus.\label{corpus-table}}
\end{table}

@@

A useful way of modelling user interfaces is to consider their
configurations as state vectors and to deploy the methods of matrix
algebra to describe transitions between these states during an
interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}.
An opposite approach is to record users' interactions with an
interface, manually coding interaction states. 
This method has been applied to usability analysis of resource
planning software for security applications~\cite{Kannampallil:2007fp}
where the transition probability matrix of these state sequences
was used to draw conclusions about the nature of this interactions.
In the creative arts, editing commands at a computer interface for the
``live coding'' of computer music has also been coded as states and
analysed using transition matrices~\cite{Swift:2014tya}, in both the
textual and musical dimensions, in order to arrive at conclusions
regarding artistic style.

>>>>

\begin{abstract}
  In this note we describe the analysis of a data archive for an
  ensemble musical performance practice on touch screens where
  transition matrix measures are able to distinguish between sessions
  of different contexts, and the internal structure of these sessions.
  Traditional methodologies for studying interfaces that support
  collaborative creativity can be time-consuming and intrusive to
  implement but our methods may provide a practical alternative for
  large scale analysis of creative interactions.

  Our archive includes 95 touch data logs of ensemble touch-screen
  interactions in concerts, rehearsals and formal HCI studies of
  groups of musicians over more than two years. The data analysis has
  been undertaken by first classifying these data into higher level
  gesture types and summarising the performers' transitions between
  these gestures. We find that two transition-matrix measure, flux,
  and entropy, can be used to distinguish different performance types
  and contexts, as well as the internal structure (beginning, middle
  and end) of the performances in our corpus.
\end{abstract}

@@

Modern computing devices and interfaces continue to present
opportunities to support new kinds of creative expression and
collaborative interaction~\cite{Resnick:2005yu}. However, it has been
acknowledged that understanding and evaluating interfaces in creative
contexts can be challenging~\cite{Shneiderman:2007qv}. In the context
of live performance, interface evaluation using questionnaires, or
ethnographic data gathering can miss the dynamic needs of artists
during collaborative interactions and may be impractical in live
performative contexts. When performers use a combuter-based artistic
interface, logged interaction data which has been collected for
curation and preservation~\cite{England:2014ys}, might also allow
quantitative investigation into these performances. In this note we
will show how an archive of collaborative performance data logs can be
subjected to a quantitative analysis based on transition probabilities
between gestural states. We will demonstrate that two measures on
transition matrices, flux and entropy, show significant differences
between different kinds of performances, and give insight into the
internal structure of these performances.

Our archive of interaction data logs has been accumulated to support a
long-term practice-led research program into simple iPad interfaces
for group music-making similar to those described in Martin et
al~\cite{Martin:2014cr}. Over a period of more than two years (April
2013--June 2015), we collected data from group music-making sessions
where the participants used our custom-built, touch-screen apps on
Apple iPad devices. Although several different apps have been included
in our corpus, they shared a common mode of free-form touch
interaction where tapping produced short sounds and swirling or
swiping produced long sounds. Apart from this commonality, the apps
featured a variety of sound palettes, different kinds of visual
feedback, and different kinds of networked interactions (between the
iPads in an ensemble and with a central server).

In all of our interaction sessions, every touch interaction---every
touch down, dragging movement and release---was time-stamped and
logged to disk for later analysis. These sessions included
formally-structured HCI experiments, where participants were
interviewed or surveyed. They also included rehearsals without an
audience and performances at arts festivals where creative goals were
paramount; in these situations surveys or interviews would have been
inappropriate and possibly disruptive. Some of the performances were
iPad-only and others used iPads together with other acoustic
percussion instruments. Although the majority of the sessions were
free-form improvisations, a number of rehearsals and performances
involved composed pieces where the musicians read and performed
percussion notation using the apps.

Representative photographs of some of these session contexts are are
shown Figure~\ref{fig:three-performance-contexts}) and other
descriptive statistics about the sessions are shown in Table
\ref{corpus-table}. In our corpus, the median length of sessions is
7m26s and the median number of participants per ensemble is four. Each
session record is labelled with metadata about the session
\emph{context} (performance, rehearsal, study), session \emph{type}
(improvisation or composition), and \emph{instrumentation}
(iPads-only, or iPads with acoustic percussion). The distribution of
recorded sessions by context and type is shown in Figure
\ref{fig:count-data}.

@@

\begin{table}
\centering
\begin{tabular}{l|llll}
\hline
            & Total & Min  & Median   & Max     \\ 
\hline
N           & 95    &      &          &         \\
%Record size & 306MB     &         &        &   \\
Length & 13H9M19S & 1M19S & 7M26S & 22M20S \\
Participants& 353   & 2    & 4        & 9        \\
Flux &   & 0.02239 & 0.1445 & 0.3246\\
Entropy &   & 0.7683 & 3.342 & 4.442\\          
\hline
\end{tabular}
\caption{
  Representative data about the corpus of musical touch-screen
  interaction sessions used in this study. These sessions were
  rehearsals, performances, or studies and
  participants improvised or performed compositions. Each record
  consisted of touch-data recorded in CSV format and was later analysed
  with a gestural classification system. Some participants were
  present in multiple sessions, however this data was not tracked in
  the corpus.\label{corpus-table}}
\end{table}

@@

A useful way of modelling user interfaces is to consider their
configurations as state vectors and to deploy the methods of matrix
algebra to describe transitions between these states during an
interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}. An
inverted approach is to record users' interactions with an interface,
manually coding interaction states  This method has been applied to
usability analysis of resource planning software for security
applications~\cite{Kannampallil:2007fp} where the transition
probability matrix of these state sequences was used to draw
conclusions about the nature of this interactions. In the creative
arts, editing commands at a computer interface for the ``live coding''
of computer music has also been coded as states and analysed using
transition matrices~\cite{Swift:2014tya}, in both the textual and
musical dimensions, in order to arrive at conclusions regarding
artistic style.
